MCMC Informed Variational Bayes

### 2017-05-07 MIVB/MIVB.tex
updates

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{MCMC Informed Variational Bayes}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\begin{document}

@@

Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$ such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:{t-1}$, given $\boldsymbol{\theta}$ and $x_t$. One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution. MCMC can have slow convergence rates and generally does not admit parallel processing or data subsampling. In this paper it is assumed that an MCMC sample of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ is available at time $T+S$, but convergence of MCMC is slow, so a posterior distribution conditioned on any additional observations of $y_t, t = T+1, \dots, T+S$ will not be available before observation of $y_{T+S+h}$.

With $p(\boldsymbol{\theta}, x_{0:T}, y_{1:T})$ the forecast distribution of p(y_{T+S+h} | y_{T})$ is available, however incorporating the information contained in realisations of $y_{T+1:T+S}$ to obtain $p(y_{T+S+h} | y_{T+S})$ is desirable, as it should provide a forecast with reduced uncertainty. The addition of $y_{T+1:T+S}$ is used to update the conditional posterior distribution of $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:T})$ to $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:{T+S})$, which can be performed with filtering techniques such as the Kalman filter with certain restrictions on the model. If these restrictions are not met, or the filtering technique is too computationally intensive to perform before $T+S+h$, then the forecaster is unable to obtain $p(y_{T+S+h} | y_{T+S})$.

An alternative to MCMC is Variational Bayes (VB), which approximates the posterior distribution with a tractable alternative denoted by $q$, optimising the parameters of this approximation so that the approximation is close by some metric to the actual posterior. A VB optimisation algorithm is often orders of magnitude faster than MCMC with the application of parallel processing and data subsampling available, so the approximate posterior $q(\boldsymbol{\theta}, x_{0:T+S} | y_{1:T+S})$ is available before $T+S+h$, so the approximate forecast distribution $q(y_{T+S+h} | y_{T+S})$ can be obtained.

The use of VB introduces an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error. This paper explores that trade-off for a Dynamic Linear Model and a Stochastic Volatility Model, two common latent state time series models, and introduces MCMC Informed Variational Bayes (MIVB), that emphasises the use of information available in MCMC samples of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T+S})$ as an alternative to standard `black box' approaches where a generic functional form is used for all models.

\begin{itemise}
\item Section Two: Variational Bayes, Optimisation and ADVI
\item Section Three: MCMC Informed Variational Bayes
\item Section Four: Simulation results for DLM, SVM
\item Section Five: Empirical SVM
\item Section Six: Discussion
\end{itemise}

@@

This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T+S} |y_{1:T+S})}\right) d\boldsymbol{\theta} dx_{1:T+S}
\end{equation}
and can alternatively be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \ln(p(y_{1:T+S})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T+S}))$.  $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{1:T+S})}{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T+S},
\end{equation}
and as $\ln(p(y_{1:T+S}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).

@@

The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and a differentiable transform $f$ such that
\begin{equation}
\label{Reparam}
\boldsymbol{\theta}, x_{1:T+S} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
\end{equation}
Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. Applying the transformation to $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ results in
\begin{equation}
\label{changevar}
q(\boldsymbol{\epsilon} = q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
\end{equation}
where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T+S}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
\begin{align}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
&= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
&= \int q(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
\end{align}

The derivative of (\ref{reparamELBO}) can be estimated by
\begin{equation}
\label{ReparamDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T+S}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)))\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
\end{equation}
where simulations of $\boldsymbol{\theta}$ and $x_{1:T+S}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $N = 1$ provides a precise enough estimate of the gradient for fast convergence.

@@

\begin{align}
y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
\end{align}
where $\epsilon_t$ and $\nu_t$ are independent $\mathcal{N}(0, 1)$ random variables. The priors used are
\begin{align}
\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
\phi &\sim \mathcal{U}(-1, 1) \\
\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
\end{align}
where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution. Conjugate priors are chosen for $\gamma, \sigma^2_y$ and $\sigma^2_x$, while the uniform prior for $\phi$ constrains $p(\phi | y_{1:T+S})$ to have zero probability mass outside of the AR(1) stationary region. $\sigma^2_y$ and $\sigma^2_x$ are log-transformed such that $\theta^* = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x), x_{0:T}\} \in \mathbb{R}^{T+5}$. $100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$. For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5,000$ and $10,000$ as this has a complexity of $O(T^2)$. MCMC uses a Gibbs sampler with $\min(5000, 10T)$ iterations, discarding the first half. The posterior means for $\boldsymbol{\theta}$ are plotted in Figure \ref{fig:DLM:ADVI}, which suggests that ADVI is inconsistent for $\phi, \sigma^2_x$ and $\sigma^2_y$. Section \label{MIVB} advocates the opposite approach to ADVI, tailoring the functional form of $q$ to the specific problem.

\begin{figure}[h]
\centering
\includegraphics{DLMAdvi}
\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$. True values for each parameter is denoted by the red line. MCMC consistently estimates of the true value, while both ADVI algorithms are only consistent for the mean parameter $\gamma$.
}
\label{fig:DLM:ADVI}
\end{figure}

A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | $\boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values. $\boldsymbol{\lambda}$ includes only the mean and variance of $\theta_i$, so the ELBO is evaluated over a grid so the stochastic nature of the gradient ascent algorithm is not a confounding factor. The maximising value of the mean of $\theta_i$ is plotted in Figure \ref{fig:DLM:Numeric}. Holding $\sigma^2_x$ and $\sigma^2_y$ fixed at their true values improves the estimation of the mean of $\phi$, but neither $\sigma^2_y$ nor $\sigma^2_x$ improves.

\begin{figure}[h]
\centering
\includegraphics{DLMNumeric}
\caption{The ELBO maximising mean for each $\theta$ parameter, holding each other $\theta$ and $x_{0:T}$ fixed. Estimation of the mean of $\phi$ improves relative to Figure \ref{fig:DLM:ADVI} but the posterior mean of $\sigma^2_y$ is consistently over-estimated while the posterior mean of $\sigma^2_x$ is consistently under-estimated.}
\label{fig:DLM:Numeric}
\end{figure}

@@

A black box VB approach such as ADVI allows implementation to be simple, as the same multivariate normal $q$ can be applied to many models, only requiring a parameter transformation. However a multivariate normal $q$ may not fit many posterior distributions well, and the ELBO may be further maximised by a different $q$. Typically, without applying the gradient ascent algorithm to many different functional forms of $q$ there is no way of determinining which distribution is optimal. However in the time-series context, MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ are available which can be utilised to provide useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$. There are two stages to this construction:
\begin{enumerate}
\item Find a well fitting functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ using the MCMC results.
\item Extrapolate the form to q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$.
\end{enumerate}

>>>>

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{MCMC Informed Variational Bayes}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\date{}
\begin{document}

@@

Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$, such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:t-1}$, given $\boldsymbol{\theta}$ and $x_t$. One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution, however these approaches can have slow convergence rates as they generally do not admit parallel processing or data subsampling to speed up computation. In this paper it is assumed that an MCMC sample of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ is available at time $T+S$, but the convergence of the required MCMC algorithm is slow, so a posterior distribution conditioned on any additional observations of $y_t, t = T+1, \dots, T+S$ will not be available before observation of $y_{T+S+h}$.

With $p(\boldsymbol{\theta}, x_{0:T}| y_{1:T})$ available, the forecast distribution of $p(y_{T+S+h} | y_{T})$ is obtainable to make an $S+h$ step ahead forecast of $y_{T+S+h}$. However, it is desirable to incorporate the information contained in realisations of $y_{T+1:T+S}$ to obtain $p(y_{T+S+h} | y_{T+S})$, which should provide a forecast with reduced uncertainty due to the smaller forecast horizon of $h$ steps. If certain model restrictions are met, $y_{T+1:T+S}$ can be included in the forecast by using filtering techniques such as the Kalman filter or particle filter. If these restrictions are not met, or the filtering technique is too computationally intensive to perform before $T+S+h$, then the forecaster is unable to obtain $p(y_{T+S+h} | y_{T+S})$.

An alternative to MCMC is Variational Bayes (VB), which approximates the posterior distribution with a tractable alternative denoted by $q$, optimising the parameters of this approximation so that the approximation is close by some metric to the actual posterior. A VB optimisation algorithm is often orders of magnitude faster than MCMC, with the application of parallel processing and data subsampling available to scale compuation to large datasets. With the use of VB the approximate posterior $q(\boldsymbol{\theta}, x_{0:T+S} | y_{1:T+S})$ is available before $T+S+h$, and thus the approximate forecast distribution $q(y_{T+S+h} | y_{T+S})$ can be used.

Variational Bayes involves an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error. This paper explores that trade-off for a Dynamic Linear Model and a Stochastic Volatility Model, two common latent state time series models, and introduces MCMC Informed Variational Bayes (MIVB), which emphasises the use of information available in MCMC samples of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T+S})$ as an alternative to standard `black box' approaches where a generic functional form is used for all models.

\begin{itemize}
\item Section Two: Variational Bayes, Optimisation and ADVI
\item Section Three: MCMC Informed Variational Bayes
\item Section Four: Simulation results for DLM, SVM
\item Section Five: Empirical SVM
\item Section Six: Discussion
\end{itemize}

@@

This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$. The KL divergence is defined by
\begin{align}
\label{KL-def}
KL[q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] &= \nonumber \\
\int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) &\ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T+S} |y_{1:T+S})}\right) d\boldsymbol{\theta} dx_{1:T+S}
\end{align}
and can alternatively be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \ln(p(y_{1:T+S})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T+S}))$.  $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{1:T+S})}{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T+S},
\end{equation}
and as $\ln(p(y_{1:T+S}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).

@@

The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and a differentiable transform $f$ such that
\begin{equation}
\label{Reparam}
\boldsymbol{\theta}, x_{1:T+S} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
\end{equation}
Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. Applying the transformation to $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ results in
\begin{equation}
\label{changevar}
q_{\epsilon}(\boldsymbol{\epsilon}) = q_{\theta, x}(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
\end{equation}
where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T+S}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
\begin{align}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\theta, x}(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\epsilon}(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q_{\epsilon}(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
&= \mathcal{L}(q(\boldsymbol{\epsilon}), \boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
\end{align}

The derivative of (\ref{reparamELBO}) can be estimated by
\begin{equation}
\label{ReparamDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\epsilon}), \boldsymbol{\lambda}^{(m)}, y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T+S}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i))) + \ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
\end{equation}
where simulations of $\boldsymbol{\theta}$ and $x_{1:T+S}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $N = 1$ provides a precise enough estimate of the gradient for fast convergence.

@@

\begin{align}
y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
\end{align}
where $\epsilon_t$ and $\nu_t$ are independent $\mathcal{N}(0, 1)$ random variables. The priors used are
\begin{align}
\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
\phi &\sim \mathcal{U}(-1, 1) \\
\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
\end{align}
where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution. Conjugate priors are chosen for $\gamma, \sigma^2_y$ and $\sigma^2_x$, while the uniform prior for $\phi$ constrains $p(\phi | y_{1:T+S})$ to have zero probability mass outside of the AR(1) stationary region. $\sigma^2_y$ and $\sigma^2_x$ are log-transformed such that $\theta^* = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x), x_{0:T}\} \in \mathbb{R}^{T+5}$. $100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2000, 5000, 10000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$. For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5000$ and $10000$ as this has a complexity of $O(T^2)$. MCMC uses a Gibbs sampler with $\min(5000, 10T)$ iterations, discarding the first half. The posterior means for $\boldsymbol{\theta}$ are plotted in Figure \ref{fig:DLM:ADVI}, which suggests that ADVI is inconsistent for $\phi, \sigma^2_x$ and $\sigma^2_y$. Section \ref{MIVB} advocates the opposite approach to ADVI, tailoring the functional form of $q$ to the specific problem.

\begin{figure}[h]
\centering
%\includegraphics{DLMAdvi}
\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2000, 5000, 10000\}$. True values for each parameter is denoted by the red line. MCMC consistently estimates of the true value, while both ADVI algorithms are only consistent for the mean parameter $\gamma$.
}
\label{fig:DLM:ADVI}
\end{figure}

A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | \boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values. $\boldsymbol{\lambda}$ includes only the mean and variance of $\theta_i$, so the ELBO is evaluated over a grid so the stochastic nature of the gradient ascent algorithm is not a confounding factor. The maximising value of the mean of $\theta_i$ is plotted in Figure \ref{fig:DLM:Numeric}. Holding $\sigma^2_x$ and $\sigma^2_y$ fixed at their true values improves the estimation of the mean of $\phi$, but neither $\sigma^2_y$ nor $\sigma^2_x$ improves.

\begin{figure}[h]
\centering
%\includegraphics{DLMNumeric}
\caption{The ELBO maximising mean for each $\theta$ parameter, holding each other $\theta$ and $x_{0:T}$ fixed. Estimation of the mean of $\phi$ improves relative to Figure \ref{fig:DLM:ADVI} but the posterior mean of $\sigma^2_y$ is consistently over-estimated while the posterior mean of $\sigma^2_x$ is consistently under-estimated.}
\label{fig:DLM:Numeric}
\end{figure}

@@

A black box VB approach such as ADVI allows implementation to be simple, as the same multivariate normal $q$ can be applied to many models, only requiring a parameter transformation. However a multivariate normal $q$ may not fit many posterior distributions well, and the ELBO may be further maximised by a different $q$. Typically, without applying the gradient ascent algorithm to many different functional forms of $q$ there is no way of determinining which distribution is optimal. However in the time-series context, MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ are available which can be utilised to provide useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$. There are two stages to this construction:
\begin{enumerate}
\item Find a well fitting functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ using the MCMC results.
\item Extrapolate the form to $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$.
\end{enumerate}
